---
title: "A Quick How-to On Labelling Bar Graphs in ggplot2"
author: "Alessandro Speranza"
date: "07/06/2021"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cash = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
# EDA packages
library(funModeling)
library(DataExplorer)
```

Dataset references can be found at [Fuel economy data from 1999 to 2008 for 38 popular models of cars](https://ggplot2.tidyverse.org/reference/mpg.html)

# Check the variables

```{r}
df_status(mpg)
```

# Create the dataset

```{r}
mpg_sum <- mpg %>%
  # just use 2008 data
  dplyr::filter(year == 2008)

# Examining the variable 'manufacturer'
str(mpg_sum$manufacturer)
# or
glimpse(mpg_sum$manufacturer)

summary(mpg_sum$manufacturer)

table(mpg_sum$manufacturer) # 15 manufacturers

# factolevels()
levels(mpg_sum$manufacturer) #NULL because the variable is still a character


# EDA
DataExplorer::plot_bar(mpg_sum$manufacturer)
```

```{r}
mpg_sum2 <- mpg %>%
  # just use 2008 data
  dplyr::filter(year == 2008) %>% 
  # turn into lumped factors with capitalized names
  dplyr::mutate(manufacturer = stringr::str_to_title(manufacturer),
                manufacturer = forcats::fct_lump(manufacturer, n = 10)
  )

# Examining the variable manufacturer (factor)
str(mpg_sum2$manufacturer)
# or
glimpse(mpg_sum2$manufacturer)

summary(mpg_sum2$manufacturer)

table(mpg_sum2$manufacturer) # 10 manufacturers + Other (11 factor levels)

# factolevels()
levels(mpg_sum2$manufacturer)


# EDA
DataExplorer::plot_bar(mpg_sum2$manufacturer)
```

```{r}
mpg_sum3 <- mpg %>%
  # just use 2008 data
  dplyr::filter(year == 2008) %>% 
  # turn into lumped factors with capitalized names
  dplyr::mutate(manufacturer = stringr::str_to_title(manufacturer),
                manufacturer = forcats::fct_lump(manufacturer, n = 10)
  ) %>% 
  # add counts
  count(manufacturer, sort = TRUE)

mpg_sum3

# factolevels()
levels(mpg_sum3$manufacturer)
```

```{r}
mpg_sum4 <- mpg %>%
  # just use 2008 data
  dplyr::filter(year == 2008) %>% 
  # turn into lumped factors with capitalized names
  dplyr::mutate(manufacturer = stringr::str_to_title(manufacturer),
                manufacturer = forcats::fct_lump(manufacturer, n = 10)
  ) %>% 
  # add counts
  count(manufacturer, sort = TRUE) %>% 
  # order factor levels by number, put "Other" to end
  dplyr::mutate(manufacturer = forcats::fct_inorder(manufacturer))

mpg_sum4

# factolevels()
levels(mpg_sum4$manufacturer)
```

```{r}
mpg_sum5 <- mpg %>%
  # just use 2008 data
  dplyr::filter(year == 2008) %>% 
  # turn into lumped factors with capitalized names
  dplyr::mutate(manufacturer = stringr::str_to_title(manufacturer),
                manufacturer = forcats::fct_lump(manufacturer, n = 10)
  ) %>% 
  # add counts
  count(manufacturer, sort = TRUE) %>% 
  # order factor levels by number, put "Other" to end
  dplyr::mutate(manufacturer = forcats::fct_rev(forcats::fct_inorder(manufacturer)))

mpg_sum5

# factolevels()
levels(mpg_sum5$manufacturer)
```

```{r}
mpg_sum6 <- mpg %>%
  # just use 2008 data
  dplyr::filter(year == 2008) %>% 
  # turn into lumped factors with capitalized names
  dplyr::mutate(manufacturer = stringr::str_to_title(manufacturer),
                manufacturer = forcats::fct_lump(manufacturer, n = 10)
  ) %>% 
  # add counts
  count(manufacturer, sort = TRUE) %>% 
  # order factor levels by number, put "Other" to end
  dplyr::mutate(manufacturer = forcats::fct_rev(forcats::fct_inorder(manufacturer)),
                manufacturer = forcats::fct_relevel(manufacturer, "Other", after = 0)
  )

mpg_sum6

# factolevels()
levels(mpg_sum6$manufacturer)

# Keep in mind that we have reversed the ordering since {ggplot2} plots factors from bottom to top when being mapped to y
```

# Data Visualization with ggplot2

```{r}
ggplot(mpg_sum6, aes(x = n, y = manufacturer)) +
  geom_col(fill = 'steelblue') +
  theme_minimal()
```

## Calculate the Percentage Values (Case 1), using a new dataset variable

```{r}
mpg_sum7 <- mpg_sum6 %>% 
  dplyr::mutate(perc = stringr::str_c(sprintf("%4.1f", n / sum(n) * 100), "%"))

mpg_sum7
```

## Calculate the Percentage Values (Case 2), using scales package -\>\> MUCH BETTER!

```{r}
mpg_sum8 <- mpg_sum6 %>% 
  dplyr::mutate(perc = scales::percent(n / sum(n), accuracy = .1, trim = FALSE))
# trim to FALSE is extremely important to right-justified all percentages

mpg_sum8
```

## Add the prepared percentage label to our bar graph with geom_text()

```{r}
ggplot(mpg_sum8, aes (x = n, y = manufacturer)) + 
  geom_col(fill = 'steelblue') +
  geom_text(aes(label = perc)) + 
  theme_minimal()
```

## Add some more description to one of the bars

```{r}
mpg_sum9 <- mpg_sum6 %>% 
  dplyr::mutate(perc = scales::percent(n / sum(n), accuracy = .1, trim = FALSE),
                perc = if_else(row_number() == 1, stringr::str_c(perc, "of all car models", sep = " "), perc))

mpg_sum9

ggplot(mpg_sum9, aes (x = n, y = manufacturer)) + 
  geom_col(fill = 'steelblue') +
  geom_text(aes(label = perc)) + 
  scale_x_continuous(limits = c(NA, 24)) +
  theme_minimal()

# or

ggplot(mpg_sum9, aes (x = n, y = manufacturer)) + 
  geom_col(fill = 'steelblue') +
  geom_text(aes(label = perc)) +
  # xlim instead of scale_x_continuous(limits) function
  xlim(NA, 24) +
  theme_minimal()
```

## Creating the same bar chart Using geom_bar()

For more info about using geom_bar(), go to [3.7 Statistical transformations](https://r4ds.had.co.nz/data-visualisation.html?q=identity#statistical-transformations)

```{r}
mpg_fct <- mpg %>% 
  dplyr::filter(year == 2008) %>%
  dplyr::mutate(
    ## add count to calculate percentages later
    total = dplyr::n(),
    ## turn into lumped factors with capitalized names
    manufacturer = stringr::str_to_title(manufacturer),
    manufacturer = forcats::fct_lump(manufacturer, n = 10),
    ## order factor levels by number, put "Other" to end
    manufacturer = forcats::fct_rev(forcats::fct_infreq(manufacturer)),
    manufacturer = forcats::fct_relevel(manufacturer, "Other", after = 0)
  )

str(mpg_fct)
levels(mpg_fct$manufacturer)

# to use geom_bar, data must have the largest granularity (they must not be grouped)
ggplot(mpg_fct, aes(x = manufacturer)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", 
            aes(label = ..count..)) +
  coord_flip() +
  theme_minimal()
```

## How to position the percentage labels inside the bars

```{r}
ggplot(mpg_sum9, aes(x = n, y = manufacturer)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = perc), 
  # make labels left-aligned)
  hjust = 1, nudge_x = -.5 ) +
  theme_minimal()
```

## How to position the percentage labels inside the bars and increase space on the right via theme(plot.margin)
```{r}
ggplot(mpg_sum9, aes(x = n, y = manufacturer)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = perc), 
  # make labels left-aligned)
  hjust = 0, nudge_x = .5 ) +
  # make sure labels doesn't get cut, part 1
  coord_cartesian(clip = "off") +
  theme_minimal() +
  ## make sure labels doesn't get cut, part 2
  theme(plot.margin = margin(r = 120))
```

## Increase space on the right via scale_x_continuous(limits)
```{r}
ggplot(mpg_sum9, aes(x = n, y = manufacturer)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = perc), 
  # make labels left-aligned)
  hjust = 0, nudge_x = .5 ) +
  # make sure labels doesn't get cut
  scale_x_continuous(limits = c(NA, 30)) +
  theme_minimal()
```

## Increase space on the right via scale_x_continuous(limits)
```{r}
ggplot(mpg_sum9, aes(x = n, y = str_c(manufacturer, perc, sep = "  "))) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = perc),
            hjust = 0, nudge_x = .1) +
  scale_x_continuous(limits = c(NA, 30)) +
  theme_minimal()

ggplot(mpg_sum9, aes(x = n, y = manufacturer)) +
  geom_col(fill = "steelblue") +
  # geom_text(aes(label = perc),
  #           hjust = 0, nudge_x = 4) +
  scale_x_continuous(limits = c(NA, 30)) +
  # scale_x_reverse() +
  # scale_y_discrete(guide = guide_axis(position = 'left')) +
  theme_minimal()

```
